---
title: "Elsevier Hybrid Open Access Analysis"
description: | 
Publishers rarly share information to whom they sent the invoice for an article, or if a fee was waived. This post presents Elsevier as an remarkable exception that provides such information in an open and machine-readable way. Analysing 70,000 hybrid open access articles published in around 1,800 Elsevier journals between 2015 and now reveals a growth of centrally covered publication fees. Nevertheless, the majority of funding sources remains unknown, raising important questions about the transparency of fee-based open access publishing.
date: Aug 29, 2019
author:
  - name: Najko Jahn 
    url: https://twitter.com/najkoja
    affiliation: State and University Library Göttingen
    affiliation_url: https://www.sub.uni-goettingen.de/
output: distill::distill_article
bibliography: literature.bib
---

```{r, echo = FALSE, message = FALSE, warning = FALSE}
knitr::opts_chunk$set(
  comment = "#>",
  collapse = TRUE,
  warning = FALSE,
  message = FALSE,
  echo = FALSE
)
options(scipen = 999, digits = 2)
knitr::knit_hooks$set(
  inline = function(x) {
    if (is.numeric(x)) {
      return(prettyNum(x, big.mark = ","))
    } else{
      return(x)
    }
  }
)
library(tidyverse)
```

In September 2018, the [cOAltion S](https://www.coalition-s.org/), a group of international research funders, announced its widely discussed [Plan S](https://www.coalition-s.org/principles-and-implementation/). According to its principles, publication fees that may arise when publishing open access should be covered by funders or research organizations. Although surveys[@Solomon_2011; @soap] suggest that many authors do not pay publication fees themselves, publishers rarely share such evidence. But also not all funders and research organizations report open access spendings publicly, resulting in an intransparent situation where policy-making and analytics lack essential data about transitioning subscription journals to open access.   <!--, despite existing attempts to crowd-source such data[@Jahn_2016]. Prominent examples sharing details about publication fee spending openly are the British [Charity Open Access Fund
](https://wellcome.ac.uk/funding/guidance/charity-open-access-fund) maintained by the Wellcome Trust or the [Open APC Initiative](https://github.com/OpenAPC/openapc-de). --> 

This blogpost presents a dataset comprising publicly available hybrid open access sponsorship information from Elsevier, a major publisher of scholarly journals. This dataset, which was created using metadata from Crossref and mining open access full-texts, serves as critical input to ongoing discussions around transition subscription journals to open access. The methods used to obtain the data not only address key challenges to discover hybrid open access articles along with funding and affiliation information with open tools and services. Elsevier's effort to make such data openly available also serves as a good practice example, potentially feeding into workflow guidelines for transformative agreements like the [ESAC guidelines](https://esac-initiative.org/about/transformative-agreements/guidelines-for-transformative-agreements/).

To demonstrate its potential, the dataset will be used to present the number and proportion of hybrid open access articles among Elsevier journals. Drawing on Elsevier's funding information, it will be also investigated whether publication fees were billed to authors or to funders that made an [agreement with Elsevier](https://www.elsevier.com/about/open-science/open-access/agreements), or if the fees were waived. Moreover, text-mined author email domains will be presented as rough approximation of the affiliation of the first resp. corresponding author, an important data point for delineating open access funding.

The resulting dataset is openly available on GitHub along with the source code.

## Methods

As a start, the [Elsevier publication fee price list](https://www.elsevier.com/__data/promis_misc/j.custom97.pdf), shared as pdf document, was used to obtain hybrid open access journals. The rOpenSci [tabulizer](https://github.com/ropensci/tabulizer) package allowed to extract data about these journals from this file. 

<aside>
  See [fetch_apc_list.R](https://github.com/njahn82/elsevier_hybrid_volume/blob/master/R/fetch_apc_list.R) how to obtain price data 
</aside>

Following the [Hybrid Open Access Journal Dashboard](https://subugoe.github.io/hybrid_oa_dashboard/about.html), an interactive analytical application from the SUB Göttingen, Crossref REST API was queried to discover open access articles published in these journals drawing on facet field counts along with the yearly article volumes for the period 2015 - 2019. After matching license URLs indicating open access articles, a second API call checked license metadata per journal. Here, using the Crossref’s REST API filters `license.url` and  `license.delay` allowed to exclude delayed open access articles. For every immediate open access article, comprehensive Crossref metadata was obtained including full-text links.

<aside>
  See [fetch_cr_md.R](https://github.com/njahn82/elsevier_hybrid_volume/blob/master/R/fetch_cr_md.R) how Crossref was called with R 
</aside>

Elsevier provides access to full-texts as `html` and `xml` document via the [Crossref Text and Data Mining Services (Crossref-TDM)](https://support.crossref.org/hc/en-us/articles/215750183-Crossref-Text-and-Data-Mining-Services). Surprisingly, the `xml` representation not only contains the full-text, but also comprehensive metadata including information about open access sponsorship shown below. 

```xml
<openaccess>1</openaccess>
<openaccessArticle>true</openaccessArticle>
<openaccessType>Full</openaccessType>
<openArchiveArticle>false</openArchiveArticle>
<openaccessSponsorName>
  Arts and Humanities Research Council
</openaccessSponsorName>
<openaccessSponsorType>FundingBody</openaccessSponsorType>
<openaccessUserLicense>
  http://creativecommons.org/licenses/by/4.0/
</openaccessUserLicense>
```
 
*Snapshot of open access metadata in Elsevier XML full. <https://api.elsevier.com/content/article/pii/S1475158518302261>*

After interfacing the Elsevier full-texts with the [crminer](https://docs.ropensci.org/crminer/) package, a client maintained by rOpenSci, the above-highlighted open access information was extracted from the `xml documents. 

<aside>
  See [R/tdm_oa_info.R](https://github.com/njahn82/elsevier_hybrid_volume/blob/master/R/tdm_oa_info.R)
</aside>

Moreover, the first author email address was parsed using pattern matching, assuming that email domains roughly indicate the affiliation of the first respective corresponding author at the time of publication. Next, the email domains was split in its parts with [urltools](https://CRAN.R-project.org/package=urltools). 


<aside>
  See [R/extract_email_domains.R](https://github.com/njahn82/elsevier_hybrid_volume/blob/master/R/extract_email_domains.R)
</aside>

## Dataset characteristics and availability

The resulting dataset comprises the following variables, and is openly shared via GitHub.


First ten rows

```{r, layout="l-body-outset", echo = TRUE}
library(rmarkdown)
hybrid_df <- readr::read_csv("data/els_hybrid_info_normalized.csv")
paged_table(head(hybrid_df, 10))
```

It must be noted, however, that open access information from Elsevier full-text was not documented at the time of writing this blogpost. 

## Results

In total, the dataset comprises `r nrow(hybrid_df)` hybrid open access articles from `r hybrid_df %>% distinct(issn) %>% nrow()` hybrid open access journals published between January 2015 and July 2019. 

### What is the uptake of hybrid open access among Elsevier journals?

Using this datasets, the share of hybrid open access articles per journal was calculated. To explore variations among journals, [Bob Rudis](https://twitter.com/hrbrmstr) [ggeconodist package](https://cinc.rud.is/web/packages/ggeconodist/index.html) was used. The package does a great job replicating a boxplot aesthetics from The Economist magazine. 

<!-- To start with, we load two data files: `els_hybrid_info_normalized.csv` contains funding information at the article-level, while `data/journal_facets.json` provides figures about the yearly publication volume for every hybrid open access journal.-->


```{r, echo = FALSE}
library(tidyverse)
library(jsonlite)
library(ggeconodist)
hybrid_df <- readr::read_csv("data/els_hybrid_info_normalized.csv")
cr_facets <- jsonlite::stream_in(file("data/journal_facets.json"), verbose = FALSE)

all_per_year <- cr_facets %>%
  unnest(year_published) %>%
  rename(year = .id, article_volume = V1) %>%
  filter(year > 2014)

oa_per_year <- hybrid_df %>%
  mutate(issued_year = as.character(issued_year)) %>%
  group_by(issn, issued_year) %>%
  summarise(n = n()) %>%
  right_join(all_per_year, by = c("issn", "issued_year" = "year")) %>%
  mutate(n = ifelse(is.na(n), 0, n)) %>%
  mutate(oa_share = n / article_volume)

perc_df <- oa_per_year %>%
  filter(n != 0) %>%
 group_by(issued_year) %>% 
  summarize(median = median(oa_share),
            p_10 = quantile(oa_share, 0.1),
            p_90 = quantile(oa_share, 0.9)) %>%
  ungroup()
a_plot <- perc_df %>%
  ggplot()+
  geom_econodist(aes(x = issued_year, ymin = p_10, median = median, ymax = p_90),
         stat = "identity", show.legend = TRUE) +
  scale_y_continuous(position = "right", limits = range(0, 0.10),
                    labels =  scales::percent) +  coord_flip() +
  labs(
    x = NULL, y = NULL,
    title = "What is the uptake of hybrid open access among Elsevier journals?",
    subtitle = "Proportion of immediate open access articles published in subscription-based journals\nJan 2015 to July 2019",
    caption = "Data source: Elsevier B.V. / Crossref"
  ) +
  theme_minimal(base_family = "Roboto") +
  theme(panel.grid.minor = element_blank()) +
  theme(axis.ticks = element_blank()) +
  theme(panel.grid.major.y = element_blank()) +
  theme(panel.border = element_blank())
grid.newpage()
left_align(a_plot, c("subtitle", "title", "caption")) %>% 
  add_econodist_legend(econodist_legend_grob(family = "Roboto"), below = "subtitle") %>% 
  grid.draw()  
```

The figure shows a slow, but steady hybrid open access uptake. The median open access proportion was around `r perc_df %>% filter(issued_year == "2019") %>% .$median %>% round(2) * 100`% in the first seven months in 2019. `r hybrid_df %>% distinct(issn) %>% nrow()` of `r cr_facets %>% distinct(issn) %>% nrow()` subscription journals from Elsevier offering hybrid open access did in fact publish at least one article under this model, corresponding to an share of `r hybrid_df %>% distinct(issn) %>% nrow() / cr_facets %>% distinct(issn) %>% nrow() * 100` %. 

### Hybrid open access facilitated by institutional agreements with Elsevier

Elsevier usually requires authors to pay a publication fee, also known as article processing charge (APC) to publish open access. Many authors make use of funding from grant agencies or academic institutions to cover such fees. To streamline this process, some funding bodies and institutions have agreed central payment options for affiliated researcher. Elsevier also provides APC waivers.

```{r}
sponsored_articles <- hybrid_df %>%
  mutate(oa_sponsor_type = ifelse(is.na(oa_sponsor_type), "unknown", oa_sponsor_type)) %>%
  mutate(fct_source = fct_lump(oa_sponsor_type, prop = 0.05)) %>%
  count(issued_year, fct_source)
all_articles <- hybrid_df %>%
  group_by(issued_year) %>%
  summarise(n = n())

ggplot(sponsored_articles, aes(x = gsub("20", "", issued_year), y = n)) +
  geom_bar(
    data = all_articles,
    aes(fill = "All hybrid open access articles"),
    color = "transparent",
    stat = "identity"
  ) +
  geom_bar(aes(fill = "by type of recipient"), color = "transparent", stat = "identity") +
  facet_wrap( ~ fct_source, nrow = 1) +
  scale_fill_manual(values = c("#b3b3b3a0", "#56B4E9"), name = "") +
  labs(x = "Year", y = "Hybrid OA Articles (Total)", title = "Elsevier Hybrid Open Access:\nTo whom did the publisher send the bill?",
       caption = "Data source: Elsevier B.V. / Crossref") +
  scale_y_continuous(labels = scales::number_format(big.mark = " ")) +
  theme_minimal(base_family="Roboto", base_size = 12) +
  theme(panel.grid.minor=element_blank()) +
  theme(axis.ticks=element_blank()) +
  theme(panel.grid.major.x=element_blank()) +
  theme(panel.border=element_blank()) +
  theme(legend.position="top", legend.justification = "right")
```

In most cases, payment notifications were send to the authors paid directly `r round(hybrid_df %>% filter(oa_sponsor_type == "Author") %>% nrow() / nrow(hybrid_df) * 100, 2)` %. Elsevier lists a funding body covering the open access publication fee for around one third of articles. 

The following interactive visualization let's you browse for funders as disclosed by Elsevier.

```{r}
library(echarts4r)
hybrid_df %>% 
  replace_na(list(oa_sponsor_name = "unknown", oa_sponsor_type = "unknown")) %>%
  count(oa_sponsor_name, oa_sponsor_type) %>%
    e_charts() %>%
    e_treemap(oa_sponsor_type, oa_sponsor_name, n, leafDepth = "1",
              name = "APC Source") %>%
    e_tooltip(trigger = "item") %>%
    e_title("Elsevier Hybrid Open Access: APC Billing Recipients")
```

Mostly British and Dutch funders sponsored hybrid open access in Elsevier journals. But also the German Federal Ministry of Education and Research (BMBF) is well represented despite the [current boycott from most universities and research organizations in Germany](https://www.nature.com/articles/d41586-018-05754-1). Since 2018, the BMBF financially supported `r hybrid_df %>% filter(grepl("BMBF", oa_sponsor_name)) %>% nrow()` hybrid open access articles that appeared in `r hybrid_df %>% filter(grepl("BMBF", oa_sponsor_name)) %>% distinct(issn) %>% nrow()` Elsevier journals according to the publisher.

### Author affiliation

In addition to funding information, email domains were parsed from Elsevier full-texts. These domains roughly indicate the affiliation of the first or of the corresponding authors, respectively, a data point used to delineate open access funding. In the following, a hierarchical, interactive treemap visualizes the distribution of the email domains. Each top-level domain can be subdivided further into domain names representing academic institutions or companies. The size of each rectangle is proportional to the number of hybrid open access articles corresponding to this domain.

```{r}
library(echarts4r)
hybrid_df %>% 
  mutate(domain = ifelse(!is.na(domain), paste(domain, suffix, sep ="."), NA)) %>%
  replace_na(list(suffix = "unknown", domain = "unknown")) %>%
  count(suffix, domain) %>%
    e_charts() %>%
    e_treemap(suffix, domain, n, leafDepth = "1",
              name = "Email Domain") %>%
    e_tooltip(trigger = "item") %>%
    e_title("Email Domains from First / Corresponding Author
")

```

## Discussion and conclusion

<!-- open data sets and methods, helpful for research about the open acces but, but alos valuable for its implementation, for instance to prepar enoegoitons between publsihers and libraires
- value of rich metadata provided by publisher to complement the existing evidence base, mostly driven funders and lirbaries
- shows that it is not a technolgical problem, but many services as well as data analytics tools are already in place (crossref, tdm) 
- more practical, although mostly paid to .., growing importance of aggrements making open access funding more transparent
- however, mostly unknown-->
